---
title: "4T"
author: "Kalyani Cauwenberghs"
date: "10/22/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## R functions revisited

Source my functions from last day
```{r}
source("http://tinyurl.com/rescale-R")
```

```{r}
rescale2(c(1:10),"barry")
```

```{r}
rescale(c(1:10),"barry")
```

Write a function to find where NA elements are in input vectors:
First make an input vector:
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
```

```{r}
is.na(x)
is.na(y)
```
Get positions of NA's
```{r}
which(is.na(x))
which(is.na(y))
```
Find where both x and y have an NA
```{r}
is.na(x)&is.na(y)
```
Taking sum tells us how many true elements in the vector.
```{r}
sum(is.na(x)&is.na(y))
```

Making a function that does the above
```{r}
bothNA<-function(x,y){
  sum(is.na(x)&is.na(y))
}
```

```{r}
bothNA(x,y)
```

But what if the vectors are of different length?
```{r}
x <- c(NA, NA, NA)
y1 <- c(1, NA, NA)
y2 <- c(1, NA, NA, NA)
both_na(x, y1)
both_na(x, y2)
```

Recycling: lengthening vectors
```{r}
x3 <- c(1, NA, NA)
x3 <- c(1, NA, NA, 1, NA, NA, 1)
y3 <- c(1, NA, NA, NA, NA, NA, NA)
```

Make a new function:
```{r}
bothNA2<-function(x,y){
  if(length(x)!=length(y)){
    stop("Passed arguments not same length.")
  }
  sum(is.na(x)&is.na(y))
}
```

```{r}
bothNA2(x,y3)
```

Write a function that returns a grade with lowest score dropped
```{r}
grade<-function(x){
  x <- replace(x, is.na(x), 0)  
  x<-sort(x)
  mean(x[2:length(x)])
}
```

Write another function like above
```{r}
grade2<-function(x){
  if(any(is.na(x))){
    stop("student is missing HW.")
  }
  sum(mean(x[-which.min(x)],ma.rm=true))/length(x)-1
}
```

Test function
```{r}
# student 1
x1<-c(100, 100, 100, 100, 100, 100, 100, 90)
# student 2
x2<-c(100, NA, 90, 90, 90, 90, 97, 80)
# student 2
x3<-c(100, NA, NA, NA, NA, NA, NA, NA)
```

```{r}
  grade(x1)
  grade(x2)
  grade(x3)
  grade2(x3)
  # now grade all students in an example class
  url <- "https://tinyurl.com/gradeinput"
  hw<-read.csv(url,row.names=1)
  apply(hw,1,grade2)
```

Explore a Bioconductor package
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Rcade")
```
Run Vignettes
```{r}
source("http://127.0.0.1:24970/library/Rcade/doc/Rcade.Rnw")
source("Rcade.R")
```














