---
title: "6Th"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Section 1: In silico docking of drugs to HIV-1 protease
We want to produce a protein-only PDB file and drug-only PDB file. 
```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
hiv <- read.pdb(file.name)
hiv
```
Q1: What is the name of the two non protein resid values in this structure? What does resid correspond to and how would you get a listing of all reside values in this structure? 

HOH and MK1


```{r}
prot <- trim.pdb(hiv, "protein")
protein <- atom.select(hiv, "protein",value=T)
lig <- trim.pdb(hiv, "ligand")
ligand <- atom.select(hiv, "ligand",value=T)
write.pdb(prot, file="1hsg_protein.pdb")
write.pdb(lig, file="1hsg_ligand.pdb")
```



Q2: Can you locate the binding site visually? Note that crystal structures normally lack hydrogen atoms, why?

No.


```{r}

```


##Section 2: Docking ligands into HIV-1 protease

```{r}
res <- read.pdb("all.pdbqt", multi=TRUE)
write.pdb(res, "results.pdb")
```


##Section 3: Exploring the conformational dynamics of proteins