---
title: "7Th"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Section 1. Differential Expression Analysis
```{r}
library(DESeq2)
```

```{r}
colData <- read.csv("GSE37704_metadata.csv", row.names=1)
countData <- read.csv("GSE37704_featurecounts.csv", row.names=1)
# Note we need to remove the odd first $length col for deSEQ2
countData <- as.matrix(countData[,-1])
head(countData)
```
Double checkthat colnames of countData without length column are the same as rownames of colData
```{r}
#check if all entries in a vector are true
all(colnames(countData)==rownames(colData))
```

Remove any genes (rows) that have 0 for all experiments (cols)
```{r}
sumcol <- countData %*% (as.matrix(rep(1,ncol(countData))))
keep_rows<-which(sumcol!=0)
countData <- as.matrix(countData[keep_rows,])
```

Perform diff. exp. analysis
```{r}
dds <- DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds <- DESeq(dds)
```
get results
```{r}
res <- results(dds)
res
plot(res$log2FoldChange,-log(res$padj))
```
add color to plot
```{r}
#first make everything gray
mycols<-rep("gray",length(res$padj))
#then make everything with extreme fold change blue
mycols[abs(res$log2FoldChange)>2]<-"blue"
plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

#install packages
```{r}
#BiocManager::install("AnnotationDbi")
#BiocManager::install("org.Hs.eg.db")
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
#can translate between all the following databases:
columns(org.Hs.eg.db)
```

```{r}
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(countData), #IDs
                    keytype="ENSEMBL", #ID format
                    column="SYMBOL", #desired format
                    multiVals="first")
res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(countData),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(countData),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")
head(res$symbol)
head(res$entrez)
head(res$name)
#write to file, ordered by p-value
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```


##Section 2. Pathway Analysis

```{r}
#BiocManager::install( c("pathview", "gage", "gageData") )

library("pathview")
library("gage")
library("gageData")
```


```{r}
library(gage)
library(gageData)

#load the data sets
#KEGG pathways 
data(kegg.sets.hs)
#signaling and metabolic pathways
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

#map our gene set to database
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
#downregulated pathways
head(keggres$less)
```

```{r  eval=FALSE}
#make one pathway infographic
pathview(gene.data=foldchanges, pathway.id="hsa04110")


#make more pathway infographics

# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")

```


##Section 3. Gene Ontology (GO)
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```



##Section 4. Reactome Analysis
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```







